A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2024-04-23 review/response.qmd

Below is a point-by-point explanation of how we have changed the paper. Reviewers' comments are in black and **\textcolor{purple}{our responses are in red}**.

# Editor and Associate Editor

Two referees, an Associate Editor (AE), and I have reviewed your paper. The two referees hold significantly different views on the paper. As you will see, the first referee's assessment is rather negative, whereas the second is more positive. As recommended by the AE, it is essential to incorporate or highlight your methodological contributions in the paper to address the concerns raised by the first referee. Therefore, I would like to extend an opportunity for revision. However, the decision for the revision will depend on whether you can adequately address the reviewers' concerns.

**\textcolor{purple}{We have made changes to the introduction and to the conclusions being more specific about the methodological contributions of this work.}**

@@

# Reviewer: 1

I commend the authors for their work in the tidy index library.  Certainly, a user friendly and unified approach to indices is useful across many fields.  My main concern about the paper is that it doesn't seem to fit within the aims and scope of JCGS.  A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software.  This paper, while certainly useful, doesn't necessarily propose new statistical methods but rather proposes a great tidy framework for handling index calculation.  As such, for JCGS, I think the authors need to highlight the statistical contributions more than the library.  However, I don't think the authors statistical contributions (even if highlighted) are sufficiently novel to warrant publication in JCGS.  As written, I think the paper is more suited for publication in, say, Journal of Statistical Software than JCGS.

**\textcolor{purple}{We are quite disappointed to see the language in this review: "A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software." If the reviewer had read the journal scope they will find "The Journal of Computational and Graphical Statistics (JCGS) presents the very latest techniques on improving and extending the use of computational and graphical methods in statistics and data analysis." The contributions that this paper makes are in computational statistics for an especially prevalent data analysis activity, that also provides potential for making better graphics in this area, also. We have added language to the introduction and to the conclusions that hopefully helps see these contributions.}**

### 2024-04-23 review/response.qmd

**\textcolor{purple}{We have made changes to the introduction and to the conclusions being more specific about the methodological contributions of this work.}**

# Reviewer: 1

I commend the authors for their work in the tidy index library.  Certainly, a user friendly and unified approach to indices is useful across many fields.  My main concern about the paper is that it doesn't seem to fit within the aims and scope of JCGS.  A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software.  This paper, while certainly useful, doesn't necessarily propose new statistical methods but rather proposes a great tidy framework for handling index calculation.  As such, for JCGS, I think the authors need to highlight the statistical contributions more than the library.  However, I don't think the authors statistical contributions (even if highlighted) are sufficiently novel to warrant publication in JCGS.  As written, I think the paper is more suited for publication in, say, Journal of Statistical Software than JCGS.

**\textcolor{purple}{We are quite disappointed to see the language in this review: ``A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software." Reading the journal scope would have informed them that ``The Journal of Computational and Graphical Statistics (JCGS) presents the very latest techniques on improving and extending the use of computational and graphical methods in statistics and data analysis." The contributions that this paper makes are in computational statistics for an especially prevalent data analysis activity, that also provides potential for making better graphics in this area. We have added language to the introduction and to the conclusions that hopefully helps see these contributions.}**

### 2024-04-23 tidyindex.qmd
changes to intro and conclusions to address the big criticism

<!-- what is the gap - why is this statistics -->
The problem is that every index is developed in its own unique way, by different researchers or organizations, and often indexes designed for the same purpose cannot easily be compared. This violates *tidy principles* [@tidyverse] which makes it difficult to investigate the nature of calculations through a principled data pipeline. It also violates statistical principles, where quantifying and understanding uncertainty are essential to deciding on a best measure or metric [@???Spiegelhalter, someone else].

@@

The paper introduces a tidy data pipeline for constructing and analyzing indexes. It has nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication.

This work addresses aspects often overlooked in index construction: uncertainty quantification and sensitivity. The pipeline supports confidence interval calculation and sensitivity analysis to be incorporated in the index construction process. It should encourage better statistical practice wherever indexes are used as critical tools in natural and social sciences.

The pipeline approach is illustrated with examples on drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI). While indexes are widely constructed and used to inform decision-making, such as in resource allocation and policy making worldwide, little effort is made to justify the choices made during index construction or to experiment with alternatives while providing means for reproducibility. This work highlights the importance of carefully constructing statistical indexes and introduces new computational methods for studying and calculating statistical indexes and their uncertainties.

There are many potential directions for the development of the work. Computationally, the tidyindex framework could be extended to support other commonly used data formats, for example, NetCDF for climate indexes. Conceptually, extending the examples to re-express additional common-practice indexes in the pipeline structure will help broader adoption, and further test that the framework can indeed accommodate any and all possible indexes.

>>>>

<!-- what is the gap - why is this statistics -->
The problem is that every index is developed in its own unique way, by different researchers or organizations, and often indexes designed for the same purpose cannot easily be compared. This echoes an issue raised in @donoho_50_2017, that different data analysts might arrive at different conclusions despite using the same data. This is especially pertinent to index use which affects important decisions such as in natural disaster prevention, economic interventions, resource allocation or human development. It is primarily due to a lack of standards in data analysis workflow. @donoho_50_2017 called for research on structuring unified workflow to address methodological variation across studies in data science. Current practices also violate statistical principles, where quantifying and understanding uncertainty are essential to deciding on a best measure or metric, for example by incorporating bootstrap confidence intervals [@efron_bootstrap_1979]. There has been considerable research in tidying up routine data analyses [XXX tidyr, dplyr, tidymodels, ..., tsibble, cubble] that turn ideas into software quickly and faithfully, as envisioned by @chambers. Index development and use needs tidying. <!--In Greater Data Science (GDS) 6: Science About Data Science, Donoho envisions future research on structuring unified workflow to address methodological variation across studies in data science. -->.

@@

The paper introduces a tidy data pipeline for constructing and analyzing indexes. It has nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication. This addresses statistical principles absent from current index definitions: uncertainty quantification and sensitivity. The tidyindex framework should encourage better statistical practice wherever indexes are used as critical tools in natural and social sciences.

Several examples are shown illustrating usage. For the drought indexes (SPI and SPEI) we showed how multiple indexes can be computed with a range of parameter choices, and compared across space and time. We showed how bootstrap confidence intervals can be readily computed and plotted to assess uncertainty about the index values, and how it may be used to make better decisions on drought declarations. The Global Gender Gap Index (GGGI) was used to illustrate how choices in dimension reduction can radically affect index values and country rankings. This also illustrates how the pipeline feeds nicely into advanced interactive graphics.

There are many potential directions for the development of the work. Computationally, the tidyindex framework could be extended to support other data formats, like NetCDF for climate indexes. Conceptually, extending the examples to re-express additional common-practice indexes in the pipeline structure will help broader adoption, and further test that the framework can indeed accommodate any and all possible indexes.

### 2024-04-23 tidyindex.tex
changes to intro and conclusions to address the big criticism

The problem is that every index is developed in its own unique way, by
different researchers or organizations, and often indexes designed for
the same purpose cannot easily be compared. This violates \emph{tidy
principles} (Wickham et al. 2019a) which makes it difficult to
investigate the nature of calculations through a principled data
pipeline. It also violates statistical principles, where quantifying and
understanding uncertainty are essential to deciding on a best measure or
metric {[}@???Spiegelhalter, someone else{]}.

>>>>

The problem is that every index is developed in its own unique way, by
different researchers or organizations, and often indexes designed for
the same purpose cannot easily be compared. This echoes an issue raised
in Donoho (2017), that different data analysts might arrive at different
conclusions despite using the same data. This is especially pertinent to
index use which affects important decisions such as in natural disaster
prevention, economic interventions, resource allocation or human
development. It is primarily due to a lack of standards in data analysis
workflow. Donoho (2017) called for research on structuring unified
workflow to address methodological variation across studies in data
science. Current practices also violate statistical principles, where
quantifying and understanding uncertainty are essential to deciding on a
best measure or metric, for example by incorporating bootstrap
confidence intervals (Efron 1979). There has been considerable research
in tidying up routine data analyses {[}XXX tidyr, dplyr, tidymodels,
\ldots, tsibble, cubble{]} that turn ideas into software quickly and
faithfully, as envisioned by (\textbf{chambers?}). Index development and
use needs tidying. .
